---
title: "User Interaction in WebGL"
author: "Duncan Murdoch"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{User Interaction in WebGL} 
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(warning = FALSE)

library(tidyverse)
library(lubridate)
library(viridis)
library(zoo)
library(janitor)
library(broom)
library(ggthemes)
library(directlabels)
library(kableExtra)
library(stargazer)
library(scales)
library(ggrepel)


colors_tableau10<-function()
{
  return(c("#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD", "#8C564B",
           "#E377C2", "#7F7F7F", "#BCBD22", "#17BECF"))
}

class_clean<-function(){
  theme_classic()+
    theme(
      plot.subtitle = element_text(color="grey10",size=rel(.7)),
      plot.title = element_text(face="bold",size=rel(.8)),
      plot.caption = element_text(color="grey50",size=rel(.5),hjust=0),
      legend.title = element_text(color="grey10",size=rel(.5)),
      legend.text = element_text(color="grey10",size=rel(.5)),
      axis.text = element_text(size=rel(1.5)),
      axis.title = element_text(size=rel(1.5)),
      #panel.grid.major = element_line(size=0,colour = "black"), 
      #plot.margin = margin(t = .1, r = .1, b = .1, l = .1,unit= "cm"),
      plot.margin = margin(t = .5, r = 1, b = .25, l = 1,unit= "cm"),
      #axis.text.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
      NULL)
}

```


## Introduction

This document describes how to use embedded Javascript to 
control a WebGL display in an HTML document.  For more 
general information, see [rgl Overview](rgl.html).

We start with two simple examples.  The next section gives 
reference information.

Consider the simple plot of the iris data.  We
insert a code chunk with label `plot3d` (which will be used below).

```{r plot3d, rgl=TRUE}
with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  type="s", col=as.numeric(Species)))
#subid <- currentSubscene3d()
```
The `rotate()` function makes use of the global `<prefix>rgl`
object.  The `knitr` WebGL support sets the prefix to the 
code chunk label, so the global is called `plot3drgl`:

    <script type="text/javascript">
    var rotate = function(angle) {
      plot3drgl.userMatrix[`r rinline("subid",
                           script=TRUE)`].rotate(angle, 0,1,0);
      plot3drgl.drawScene();
    }
    </script>



# Introduction

```{r, making_graphs,echo=FALSE,include=TRUE}
U <- function(x,y) (x*y)
utility_map(5,5,U,u_vals = c(2,4,6,8,10),x_shift = -.35,y_shift = .25)

```
---
# Introduction 2

```{r testgl,echo=FALSE,include=TRUE}
library(rgl)
x <- sort(rnorm(1000))
y <- rnorm(1000)
z <- rnorm(1000) + atan2(x,y)
plot3d(x, y, z, col=rainbow(1000))
rglwidget()
```
---

# test

```{r testrgl,rgl=TRUE}
with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  type="s", col=as.numeric(Species)))
```